Formalization of Typed Lua

### 2015-06-06 doc/papers/paper/paper.tex
Removing dangling pointers to thesis chapters

\section{Introduction}

@@

The early error detection and better tools for structuring
programs are two advantages of statically typed languages that
lead programmers to migrate their code from a dynamically
typed to a statically typed language when their scripts
evolve into complex programs~\cite{tobin-hochstadt2006ims}.
This migration from dynamic to static usually involves
different languages, which distinct syntax and semantics,
requiring a complexe rewrite of existing programs instead of
incremental evolution.

@@

Typed Lua is an optional type system for
Lua that is rich enough to preserve some of the idioms
that Lua programmers are already familiar with, while
helping programmers structure large Lua programs and
catch errors at compile-time. An informal
presentation of the design of Typed Lua has been
published~\cite{dylatypedlua}.

@@

The rest of the paper is organized as follows: Section~2
presents the methodology and results of our survey of Lua
idioms; Section 3 presents the types of Typed Lua's
type system, and their subtyping and consistent-subtyping
relationd; Section 4 presents the typing rules for
evolution of table types; Section 5 presents the typing rules
for destructuring unions of tuples into projection types
and flow typing on these projections; Section 6 reviews
related work; Section 7 presents our conclusions, and
outlines our plans for future work.

\section{Statistics about the usage of Lua}

@@

\section{Typing Rules}

In the previous chapter we presented an informal overview of Typed Lua.
We showed that programmers can use Typed Lua to combine static and dynamic
typing in the same code, and it allows them to incrementally migrate from
dynamic to static typing.
This is a benefit to programmers that use dynamically typed languages
to build large applications, as static types detect many bugs
during the development phase, and also provide better documentation.

In this chapter we present the abstract syntax of Typed Lua types,
the subtyping rules, and the most interesting typing rules.
Besides its practical contributions, Typed Lua also has some interesting
contributions to the field of optional type systems for scripting
languages.
They are novel type system features that let Typed Lua cover several Lua idioms
and features, such as refinement of tables, multiple assignment, and multiple return values.

@@

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types,
filter types, and projection types.
Second-level types include tuple types and unions of tuple types.
Tuple types include the type $\Void$, variadic types, and pair types.
Types are ordered by a subtype relationship that we introduce
in Section \ref{sec:subtyping}, so Lua values may belong to
several distinct types.

@@

The type $\Value$ is the top type, which represents any Lua value.
In Section \ref{sec:rules} we will show that this type,
along with variadic types, helps the type system to drop extra values
on assignments and function calls, thus preserving the
semantics of Lua in these cases.

Typed Lua uses the type $\Self$ to represent the \emph{receiver}
in object-oriented method definitions and method calls.
As we mentioned in Section \ref{sec:oop}, we need the type
$\Self$ to prevent programs from indexing a method without
calling it with the correct receiver.

@@

Any table type has to be \emph{well-formed}.
Informally, a table type is well-formed if key types do not overlap.
In Section \ref{sec:rules} we formalize the definition of well-formed table types.
We delay the proper formalization of well-formed table types because we use
consistent-subtyping in this formalization.

Recursive types have the form $\mu x.T$,
where $T$ is a first-level type that $x$ represents.
For instance, $\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}$
is a type for singly-linked lists of integers.
In Section \ref{sec:alias} we mentioned that we can use the following
interface declaration as an alias to this type:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we mentioned in Section \ref{sec:unions}.
This means that \texttt{type} predicates use filter types of the form
$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
$T_{2}$ is the discriminated type.

Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
In Section \ref{sec:fap} we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear in the right-hand side of the declaration of local variables,
as we mentioned in Section \ref{sec:unions}.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

Typed Lua includes the dynamic type $\Any$ for allowing programmers
to mix static and dynamic typing.

@@

Our type system uses subtyping \cite{cardelli1984smi,abadi1996to} to order
types and consistent-subtyping \cite{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We explain the subtyping and consistent-subtyping rules throughout this section.
However, we focus the discussion on the definition of subtyping because,
as we mentioned in Section \ref{sec:gradual}, we can combine the
consistency and subtyping relations to achieve consistent-subtyping.
The differences between subtyping and consistent-subtyping are the way
they handle the dynamic type, and the fact that subtyping is transitive,
but consistent-subtyping is not.

@@

The type $\Void$ is a subtype of itself and of a variadic type:
\[
\begin{array}{c}
\mylabel{S-VOID2}\\
\senv \vdash \Void \subtype T{*}
\end{array}
\]

A variadic type $T{*}$ represents a sequence of values of type
$T \cup \Nil$, and the rule \textsc{S-VOID2} handles the case where
a given sequence is empty.

@@

Note that the case where both tuple types end with the type $\Void$ does
not require any special rule.
In the next section we will show that we use the subtyping rules for variadic types,
along with the types $\Value$ and $\Nil$, to make our type system reflect
the semantics of Lua on discarding extra parameters and
replacing missing parameters.

@@

Back to the subtyping rules between first-level types,
the subtyping rule among a \emph{fixed} or \emph{closed}
table type and another \emph{closed} table type resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..n \; \exists j \in 1..m \\
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}} \; m \ge n
\end{array}
\]

@@

There is a limitation on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
If the table constructor had a \emph{closed} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describes a more general type.
For instance,
\begin{verbatim}
    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
\end{verbatim}
would not type check, as the type of the table constructor would not
be a subtype of the type in the annotation.
More precisely,
\[
\{``x":1, ``y":2\}_{closed} \not\subtype
\{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]

@@

Using \emph{unique} table types to represent the type of the table
constructor allows our type system to type check the previous example.
More precisely,
\[
\{``x":1, ``y":2\}_{unique} \subtype \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\]

Even though we allow width subtyping between \emph{unique} and \emph{closed}
table types, we do not allow it among \emph{unique} and other table types
because it would violate our definition of these other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}}
\end{array}
\]

@@

However, the rule that handles subtyping among \emph{open} and
\emph{open} or \emph{fixed} table types does not allow width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}}
\end{array}
\]

@@

In the next section we will show in more detail how our type system
uses these tags to handle the refinement of table types.

@@

As an example, the rule \textsc{S-UNFOLDR} allows our type system to
type check the function \texttt{insert} from Section \ref{sec:alias}:
\begin{verbatim}
    local function insert (e:Element?, v:integer):Element
      return { info = v, next = e }
    end
\end{verbatim}
that is, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
the type of the table constructor is a subtype of \texttt{Element}:
\begin{align*}
\{ & ``info":\Integer, ``next":\mu x.\{ ``info":\Integer, \\
   & ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
\end{align*}

@@

\subsection{Typing rules}

In this section we use a reduced core of Typed Lua to present the
most interesting rules of our type system.
These rules type check multiple assignment, table refinement,
and overloading on the return type of functions.
Appendix \ref{app:rules} presents the full set of typing rules.

@@

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
We use two different environments to handle this feature.
The first environment is the type environment $\env$ that maps
variables to first-level types.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
The second environment is the projection environment $\penv$ that
maps projection variables to second-level types.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$
with the projection variable $x$ that maps to type $S$.
In Section \ref{sec:fap} we will show how our type system uses the
projection environment $\penv$ for handling projection types,
and also for projecting unions of second-level types into
unions of first-level types.

@@

\subsubsection{Tables and refinement}

@@

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{m} = merge(\env_{1}, ..., \env_{n+1}) 
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{m}}
\end{array}
\]

Back to our example, the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}$ through rule \textsc{T-CONSTRUCTOR2}.

@@

After presenting some typing rules of the table constructor,
we start the discussion of the rules that define the most
unusual feature of our type system: the refinement of table types.
The first kind of refinement allows programmers to add new
fields to \emph{unique} or \emph{open} table types through
field assignment.
For instance, in Section \ref{sec:tables} we presented the
following example:
\begin{verbatim}

    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

We can translate this example to our reduced core as follows:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; person = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$person[``firstname"] \; {<}\String{>} = ``Lou";$}\\
& \multicolumn{1}{l}{$person[``lastname"] \; {<}\String{>} = ``Reed"$}
\end{tabular}
\end{center}

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-REFINE} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
       V = close(T)
       \end{array}}
      {\env_{1}, \penv \vdash id[k] {<}T{>}:V, \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
\end{array}
\]

@@

As an example,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``y"] \; {<}\Integer{>} = 1
\]
uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
Therefore, after these field assignments $\string_ENV$ has type
$\{``x":\String, ``y":\Integer\}_{open}$.

We do not allow the refinement of table types to add a field if it is
already present in the table's type.
For instance,
\[
\string_ENV[``x"] \; {<}\String{>} = ``foo" \;;\; \string_ENV[``x"] \; {<}\Integer{>} = 1
\]
does not type check, as we are trying to add $``x"$ twice.

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
As we mentioned in Sections \ref{sec:oop} and \ref{sec:subtyping},
we need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a top-level return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior in
a not deterministic way, as it makes easier the presentation of this behavior.

@@

Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

>>>>

\section{Introduction}
\label{sec:intro}

@@

The early error detection and better tools for structuring
programs are two advantages of statically typed languages that
lead programmers to migrate their code from a dynamically
typed to a statically typed language when their scripts
evolve into complex programs~\cite{tobin-hochstadt2006ims}.
This migration from dynamic to static typing usually involves
different languages that have distinct syntax and semantics,
requiring a complexe rewrite of existing programs instead of
incremental evolution.

@@

Typed Lua is an optional type system for
Lua that is rich enough to preserve some of the idioms
that Lua programmers are already familiar with, while
helping programmers structure large Lua programs and
catch errors at compile-time. An informal
presentation of the design of Typed Lua has been
published~\cite{maidl2014tl}.

@@

The rest of the paper is organized as follows:
Section \ref{sec:statistics} presents the methodology and results of
our survey of Lua idioms;
Section \ref{sec:types} presents the types of Typed Lua's
type system, and their subtyping and consistent-subtyping relations;
Section \ref{sec:tables} presents the typing rules for evolution of table types;
Section \ref{sec:projections} presents the typing rules for
destructuring unions of tuples into projection types and
flow typing on these projections;
Section \ref{sec:related} reviews related work;
Section \ref{sec:conclusion} presents our conclusions,
and outlines our plans for future work.

\section{Statistics about the usage of Lua}
\label{sec:statistics}

@@

\section{Types}
\label{sec:types}

In this section we present the abstract syntax of Typed Lua types
and their subtyping rules.

%In the previous chapter we presented an informal overview of Typed Lua.
%We showed that programmers can use Typed Lua to combine static and dynamic
%typing in the same code, and it allows them to incrementally migrate from
%dynamic to static typing.
%This is a benefit to programmers that use dynamically typed languages
%to build large applications, as static types detect many bugs
%during the development phase, and also provide better documentation.

%In this chapter we present the abstract syntax of Typed Lua types,
%the subtyping rules, and the most interesting typing rules.
%Besides its practical contributions, Typed Lua also has some interesting
%contributions to the field of optional type systems for scripting
%languages.
%They are novel type system features that let Typed Lua cover several Lua idioms
%and features, such as refinement of tables, multiple assignment, and multiple return values.

@@

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types.
Typed Lua splits types into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.
First-level types include literal types, base types, the type $\Nil$,
the top type $\Value$, the dynamic type $\Any$, the type $\Self$,
union types, function types, table types, recursive types,
filter types, and projection types.
Second-level types include tuple types and unions of tuple types.
Tuple types include the type $\Void$, variadic types, and pair types.
Types are ordered by a subtype relationship that we introduce
in the next section, so Lua values may belong to several distinct types.

@@

The type $\Value$ is the top type, which represents any Lua value.
This type along with variadic types help the type system to drop
extra values on assignments and function calls, thus preserving the
semantics of Lua in these cases.

Typed Lua uses the type $\Self$ to represent the \emph{receiver}
in object-oriented method definitions and method calls.
As we mentioned in our informal overview of Typed Lua \cite{maidl2014tl},
we need the type $\Self$ to prevent programs from indexing a method without
calling it with the correct receiver.

@@

Any table type has to be \emph{well-formed}.
Informally, a table type is well-formed if key types do not overlap.
In Section \ref{sec:tables} we formalize the definition of well-formed table types.
We delay the proper formalization of well-formed table types because we use
consistent-subtyping in this formalization.

Recursive types have the form $\mu x.T$,
where $T$ is a first-level type that $x$ represents.
As an example,
\[
\mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
\]
is a type for singly-linked lists of integers.
In Typed Lua, we can use the following interface declaration
as an alias to this type:
\begin{verbatim}
    local interface Element
      info:integer
      next:Element?
    end
\end{verbatim}

Typed Lua includes filter types as a way to discriminate the type of local
variables inside conditions.
Our type system uses filter types to formalize the \texttt{type} predicates
that we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
This means that \texttt{type} predicates use filter types of the form
$\phi(T_{1},T_{2})$ to discriminate local variables that are bound to
union types.
In a filter type $\phi(T_{1},T_{2})$, $T_{1}$ is the original type and
$T_{2}$ is the discriminated type.

Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types.
In Section \ref{sec:projections} we will show in more detail how our type system
uses them as a mechanism for handling unions of tuple types,
when they appear in the right-hand side of the declaration of local variables,
as we introduced in our previous design of Typed Lua \cite{maidl2014tl}.
We also show how this feature allows our type system to constrain
the type of a local variable that depends on the type of another local variable.

Typed Lua includes the dynamic type $\Any$ for allowing programmers
to mix static and dynamic typing.
There is a subtle difference between the dynamic type $\Any$ and the top type $\Value$.
Although both types mean that they accept a value of any other type,
the type $\Value$ is not a good option for handling the
interaction between dynamically typed and statically typed code.
Gradual typing uses the dynamic type $\Any$ to identify
where it should insert run-time checks for asserting that dynamically
typed code does not violate statically typed code.
Typed Lua also uses the dynamic type $\Any$ in this sense,
though it is an optional type system.
More precisely, we use $\Any$ instead of $\Value$ to allow
programmers blending dynamic and static typing because we use the
consistent-subtyping relation to formalize our optional type system,
as it is a first step to switch Typed Lua from optional typing to
gradual typing in the future.

@@

Our type system uses subtyping \cite{cardelli1984smi,abadi1996to} to order
types and consistent-subtyping \cite{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We explain the subtyping and consistent-subtyping rules throughout this section.
However, we focus the discussion on the definition of subtyping because
we can combine the consistency and subtyping relations to achieve
consistent-subtyping \cite{siek2007objects,siek2013mutable}.
The differences between consistent-subtyping and subtyping are the way
they handle the dynamic type, and the fact that subtyping is transitive,
but consistent-subtyping is not.

@@

The type $\Void$ is a subtype of itself and of a variadic type:
\[
\begin{array}{c}
\mylabel{S-VOID}\\
\senv \vdash \Void \subtype T{*}
\end{array}
\]

A variadic type $T{*}$ represents a sequence of values of type
$T \cup \Nil$, and the rule \textsc{S-VOID} handles the case where
a given sequence is empty.

@@

Note that the case where both tuple types end with the type $\Void$ does
not require any special rule.
We use the subtyping rules for variadic types along with the types
$\Value$ and $\Nil$ to make our type system reflect the semantics of
Lua on discarding extra parameters and replacing missing parameters.

@@

Back to the subtyping rules between first-level types,
the subtyping rule among a \emph{fixed} or \emph{closed}
table type and another \emph{closed} table type resembles the
standard subtyping rule between records:
\[
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..n \; \exists j \in 1..m \\
       \senv \vdash K_{j} \subtype K_{i}' \;\;\;
       \senv \vdash K_{i}' \subtype K_{j} \;\;\;
       \senv \vdash V_{j} \subtype_{c} V_{i}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
       \end{array}} \; m \ge n
\end{array}
\]

@@

There is a limitation on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
If the table constructor had a \emph{closed} table type, then
programmers would not be able to use it to initialize a variable with
a table type that describes a more general type.
For instance,
\begin{verbatim}
    local t:{ "x":integer,
              "y":integer? } = { x = 1, y = 2 }
\end{verbatim}
would not type check, as the type of the table constructor would not
be a subtype of the type in the annotation.
More precisely,
\begin{align*}
& \{``x":1, ``y":2\}_{closed} \not\subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

@@

Using \emph{unique} table types to represent the type of the table
constructor allows our type system to type check the previous example.
More precisely,
\begin{align*}
& \{``x":1, ``y":2\}_{unique} \subtype \\
& \{``x":\Integer, ``y":\Integer \cup \Nil\}_{closed}
\end{align*}

Even though we allow width subtyping between \emph{unique} and \emph{closed}
table types, we do not allow it among \emph{unique} and other table types
because it would violate our definition of these other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
       \end{array}}
\end{array}
\]

@@

However, the rule that handles subtyping among \emph{open} and
\emph{open} or \emph{fixed} table types does not allow width subtyping:
\[
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
       \end{array}}
      {\begin{array}{c}
       \senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype\\
                    \{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}
       \end{array}}
\end{array}
\]

@@

In Section \ref{sec:tables} we will show in more detail how our type system
uses these tags to handle the refinement of table types.

@@

As an example, the rule \textsc{S-UNFOLDR} allows our type system to
type check a function that inserts an element in the begining of
a singly-linked list:
\begin{verbatim}
    local function insert (e:Element?,
                           v:integer):Element
      return { info = v, next = e }
    end
\end{verbatim}
that is, the type checker uses the rule \textsc{S-UNFOLDR} to verify whether
the type of the table constructor is a subtype of \texttt{Element}:
\begin{align*}
\{ & ``info":\Integer, ``next":\mu x.\{ ``info":\Integer, \\
   & ``next":x \;\cup\; \Nil\}_{closed} \cup \Nil \}_{unique} \subtype \\
& \mu x.\{``info":\Integer, ``next":x \;\cup\; \Nil\}_{closed}
\end{align*}

@@

\section{Typing rules}
\label{sec:rules}

In this section we use a reduced core of Typed Lua to present
the typing rules that type check the novel features of our type system.
These rules type check table refinement and
overloading on the return type of functions.

@@

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
We use two different environments to handle this feature.
The first environment is the type environment $\env$ that maps
variables to first-level types.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
The second environment is the projection environment $\penv$ that
maps projection variables to second-level types.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$
with the projection variable $x$ that maps to type $S$.
In Section \ref{sec:projections} we will show how our type system uses the
projection environment $\penv$ for handling projection types,
and also for projecting unions of second-level types into
unions of first-level types.

@@

\subsection{Tables and refinement}
\label{sec:tables}

@@

The rule \textsc{T-CONSTRUCTOR2} uses these steps to type check a
table constructor with a non-empty list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash ([e_{1}] = e_{2})_{i}:(K_{i},V_{i}), \env_{i+1} \\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T) \;\;\;
       n = |\;\overline{[e_{1}] = e_{2}}\;| \\
       \env_{f} = merge(\env_{1}, ..., \env_{n+1})
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{f}}
\end{array}
\]

Back to our example, the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ has type
$\{1:``x", 2:``y", 3:``z"\}_{unique}$ through rule \textsc{T-CONSTRUCTOR2}.

@@

After presenting some typing rules of the table constructor,
we start the discussion of the rules that define the most
unusual feature of our type system: the refinement of table types.
The first kind of refinement allows programmers to add new
fields to \emph{unique} or \emph{open} table types through
field assignment.
As an example, we can translate
\begin{verbatim}
    local person = {}
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}
to our reduced core as follows:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; person = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$person[``firstname"] \; {<}\String{>} = ``Lou";$}\\
& \multicolumn{1}{l}{$person[``lastname"] \; {<}\String{>} = ``Reed"$}
\end{tabular}
\end{center}

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-REFINE} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
       V = close(T)
       \end{array}}
      {\begin{array}{l}
       \env_{1}, \penv \vdash id[k] {<}T{>}:V, \\
       \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]
       \end{array}}
\end{array}
\]

@@

As an example,
\begin{align*}
& \string_ENV[``x"] \; {<}\String{>} = ``foo" \;; \\
& \string_ENV[``y"] \; {<}\Integer{>} = 1
\end{align*}
uses field assignment to add fields $``x"$ and $``y"$ to $\string_ENV$.
Therefore, after these field assignments $\string_ENV$ has type
$\{``x":\String, ``y":\Integer\}_{open}$.

We do not allow the refinement of table types to add a field if it is
already present in the table's type.
For instance,
\begin{align*}
& \string_ENV[``x"] \; {<}\String{>} = ``foo" \;; \\
& \string_ENV[``x"] \; {<}\Integer{>} = 1
\end{align*}
does not type check, as we are trying to add $``x"$ twice.

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
We need \emph{fixed} table types to type classes in object-oriented programming.
In the implementation we fix the aliasing of \emph{unique} and \emph{open}
table types that appear in a top-level return statement, and in other cases we
close the aliasing of \emph{unique} and \emph{open} table types.
However, in the formalization we chose to define this behavior in
a not deterministic way, as it makes easier the presentation of this behavior.

@@

The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this non-deterministic behavior.
Rule \textsc{T-IDREAD1} uses the auxiliary function \emph{close} to
produce a \emph{closed} alias.
It also uses the auxiliary function \emph{open} to change the type of
the original reference from \emph{unique} to \emph{open},
because aliasing an \emph{unique} table type while keeping the original
reference \emph{unique} can be unsafe.
Rule \textsc{T-IDREAD2} uses the auxiliary function \emph{fix} to
produce a \emph{fixed} alias.
It also uses \emph{fix} to change the type of the original reference
to \emph{fixed}, because a \emph{fixed} table type does not allow
width subtyping.
We define these rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:close(T_{2}), \env_{1}[id \mapsto open(T_{1})]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env_{1}, \penv \vdash id:fix(T_{2}), \env_{1}[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

Both rules use the auxiliary function \emph{read} because they may be
accessing an identifier that is bound to a filter or projection type.
As we presented in our previous design of Typed Lua \cite{maidl2014tl},
our type system includes a small set of \texttt{type} predicates that allow
programmers to discriminate union types, and our type system uses filter and
projection types in the definition of these predicates to handle the discrimination
of unions types.
While filter types discriminate unions of first-level types, projection
types discriminate unions of second-level types and project unions of
second-level types into unions of first-level types.
We can define \emph{read} as follows:
\begin{align*}
read(\penv, \phi(T_{1},T_{2})) & = T_{2}\\
read(\penv, \pi_{i}^{x}) & = proj(\penv(x), i)\\
read(\penv, T) & = T
\end{align*}

The function \emph{read} uses the auxiliary function \emph{proj}
to project a union of first-level types, based on an union of
second-level types and an index from a projection type.
In Section \ref{sec:projections} we will discuss how our type system uses
projection types to handle overloaded return types.
We can define \emph{proj} as follows:
\begin{align*}
proj(T_{1} \times ... \times T_{n} \times T{*}, i) & =  T_{i} \;\;\; \text{if $i <= n$}\\
proj(S_{1} \sqcup S_{2}, i) & = proj(S_{1}, i) \cup proj(S_{2}, i)
\end{align*}
